Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Final submission to ITP

\acknowledgements{I want to thank \dots}%optional

@@

Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml
implementation, with
\ocaml{map (fun x -> x * 2)}.
A lot of program calculations have this pattern: start from a simple
specification, e.g.\ $\mathsf{map}\;(\lambda x.\;x \times 2) \circ
\mathsf{sort}$, and use program equivalences and algebraic laws to rewrite it to
an optimised version. We will revisit a similar example in
Section~\ref{sec:sorting}.

>>>>

% \acknowledgements{I want to thank \dots}
% optional

@@

Suppose, for example, that we want to write a program that sorts a list of
integers and multiplies them by 2 at the same time. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}

This program is the same as composing a \emph{quicksort} OCaml implementation,
with the function \ocaml{map (fun x -> x * 2)} which is a known program
optimisation called \emph{fusion}.  A lot of program calculations have this
pattern: start from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version. We will
revisit a similar example in Section~\ref{sec:sorting}.
